---
title: "HaozeZhu_HW3"
author: "Haoze"
date: "5/20/2021"
output: pdf_document
---

```{r setup}
##loading library

library(tidyverse) 
library(ROCR) 
library(tree) 
library(maptree) 
library(class) 
library(lattice) 
library(ggridges) 
library(superheat)
## loading the drug dataset and transform variales
drug_use <- read_csv('/Users/walterzhu/Rstudio/PSTAT131/PSTAT131_HW3/drug.csv',
col_names = c('ID','Age','Gender','Education','Country','Ethnicity', 'Nscore','Escore','Oscore','Ascore','Cscore','Impulsive', 'SS','Alcohol','Amphet','Amyl','Benzos','Caff','Cannabis',
'Choc','Coke','Crack','Ecstasy','Heroin','Ketamine', 'Legalh','LSD','Meth','Mushrooms','Nicotine','Semer','VSA'))

drug_use <- drug_use %>% mutate_at(as.ordered, .vars=vars(Alcohol:VSA)) 
drug_use <- drug_use %>%
mutate(Gender = factor(Gender, labels=c("Male", "Female"))) %>% mutate(Ethnicity = factor(Ethnicity, labels=c("Black", "Asian", "White", "Mixed:White/Black", "Other", "Mixed:White/Asian", "Mixed:Black/Asian"))) %>% mutate(Country = factor(Country, labels=c("Australia", "Canada", "New Zealand", "Other", "Ireland", "UK", "USA")))
```

##1. Logistic regression for drug use prediction
###a
```{r cars}
## create a new factor response variable recent_cannabis_use

drug_use = drug_use %>% mutate(recent_cannabis_use=ifelse(Cannabis >= "CL3","Yes","No"))%>% mutate(recent_cannabis_use = factor(recent_cannabis_use,levels=c("No", "Yes")))

```

###b
```{r}
##create a new tibble
drug_use_subset <- drug_use %>% select(Age:SS, recent_cannabis_use)
##Split drug_use_subset into a training data set and a test data set
set.seed(1) 
training.indices = sample(1:nrow(drug_use_subset), 1500)
drug_use_train = drug_use_subset[training.indices,]
drug_use_test = drug_use_subset[-training.indices,]
dim(drug_use_train)
dim(drug_use_test)
```
###c
```{r}
##Fit a logistic regression
glm.fit = glm(recent_cannabis_use ~ ., data=drug_use_train, family=binomial) 
# Summarize the logistic regression model 
summary(glm.fit)
```
##2. Decision tree models of drug use

```{r}
## Construct a decision tree
tree_parameters = tree.control(nobs=nrow(drug_use_train), minsize=10, mindev=1e-3)
drugtree = tree(recent_cannabis_use~.,control = tree_parameters,data = drug_use_train)
```

###a Use 10-fold CV to select the a tree
```{r}
# Set random seed 
set.seed(1)
cv = cv.tree(drugtree,FUN=prune.misclass, K=10)
best.size.cv = 6
plot(cv$size , cv$dev, type="b", 
     xlab = "Number of leaves, \'best\'", ylab = "CV Misclassification Error",
     col = "red", main="CV")
abline(v=best.size.cv, lty=2)
```

###b Prune the tree
```{r}
# prune the original tree using the best size in a
drugtree.pruned = prune.misclass(drugtree, best=best.size.cv)
draw.tree(drugtree.pruned, nodeinfo=TRUE,cex = 0.5)
```
The variable 'Country' is split ﬁrst in this decision tree.

###c Compute and print the confusion matrix for the test data
```{r}
# Predict on test set 
predictions = predict(drugtree.pruned, drug_use_test, type="class")
# get the true response of the test data
truth = drug_use_test$recent_cannabis_use
# Obtain confusion matrix
confusion_matrix = table(truth, predictions)
confusion_matrix

```
```{r}
# get true positive rate
true_positive_rate = confusion_matrix[2,2]/sum(confusion_matrix[,2])
true_positive_rate
```
```{r}
# get false positive rate
false_positive_rate = confusion_matrix[1,2]/sum(confusion_matrix[,2])
false_positive_rate
```

##3.Model Comparison
###a
```{r}
# get predition from logistic regression model using test data
prob_log_testing = predict(glm.fit,drug_use_test,type="response")
pred_log = prediction(prob_log_testing, truth)
#calculate the True Positive Rate and False Positive Rate by performance()
perf_log = performance(pred_log, measure="tpr", x.measure="fpr")
# for the decision tree model
prob_tree_testing = predict(drugtree.pruned,newdata=drug_use_test,type="prob")
pred_tree = prediction(predictions,truth)
perf_tree = performance(pred_tree, measure="tpr", x.measure="fpr")

plot(perf_log, col=2, lwd=3, main="ROC curve") 
abline(0,1)
```
###b
```{r}
##Compute the AUC for logistic regression model
auc_log = performance(pred_log, "auc")@y.values
auc_log
```
```{r}
##Compute the AUC for decision tree model

```

##4. Clustering and dimension reduction for gene expression data
```{r,results='hide'}
## read leukemia_data
leukemia_data <- read_csv("/Users/walterzhu/Rstudio/PSTAT131/PSTAT131_HW3/leukemia_data.csv")
```

###a
```{r}
##Convert the Type column to factor
leukemia_data = leukemia_data %>% mutate(Type = factor(Type))
##the number of patients with each leukemia subtype
table(leukemia_data$Type)
```
Subtype "BCR-ABL" occurs the least in this data

###b
```{r}
## exclude the Type column
leukemia_data_wt = leukemia_data %>% select(-Type)
pr.out=prcomp(leukemia_data_wt, scale=TRUE, center = TRUE)
## get the variance explained by each principal component
pr.var=pr.out$sdev ^2
pve=pr.var/sum(pr.var)
## get cumulative pve
cumulative_pve = cumsum(pve)
## plot PVE and cumulative pve
par(mfrow=c(1, 2))
plot(pve, type="l", lwd=3)
plot(cumulative_pve, type="l", lwd=3)

```
###c
```{r}
## set color
rainbow_colors <- rainbow(7) 
plot_colors <- rainbow_colors[leukemia_data$Type]
plot(pr.out$x[,1],pr.out$x[,2],xlab="PC1", ylab="PC2",cex=0.5)
text(pr.out$x[,1],pr.out$x[,2],labels = leukemia_data$Type,col=plot_colors,cex=0.6)
```
T-ALL is most clearly separated from the others along the PC1 axis
```{r}
sorted <- sort(abs(pr.out$x[,1]),decreasing = TRUE)
head(pr.out$x,1)
```
###f
```{r}
## load dendextend library
library(dendextend)
##subsetting to include only rows for which Type is either T-ALL, TEL-AML1, or Hyperdip50
leukemia_subset_1 <- filter(leukemia_data, Type == c("T-ALL","TEL-AML1","Hyperdip50"))
## exclude the first column Type
leukemia_subset = leukemia_subset_1 %>% select(-Type)
## calculate the distance matrix
subset_dist = dist(leukemia_subset)
set.seed(1)
## Hierarchical Clustering using complete linkage
drug.hclust = hclust(subset_dist)
```

```{r}
## first plot
x = as.dendrogram(drug.hclust)
x %>% set_labels(leukemia_subset_1$Type) %>% set("labels_col",k=3) %>% set("branches_k_color", k = 3) %>%
  set("labels_cex", 0.3) %>% plot(main='Three Groups for hclust',horiz = TRUE)
```
```{r}
## second plot
y = as.dendrogram(drug.hclust)
y %>% set_labels(leukemia_subset_1$Type) %>% set("labels_col",k=5) %>% set("branches_k_color", k = 5) %>%
  set("labels_cex", 0.3) %>% plot(main='Five Groups for hclust',horiz = TRUE)
```

