---
title: "131 HW3"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ROCR)
library(tree)
library(maptree)
library(class)
library(lattice)
library(ggridges)
library(superheat)
```

```{r}
drug_use <- read_csv('drug.csv',
col_names = c('ID','Age','Gender','Education','Country','Ethnicity', 'Nscore','Escore','Oscore','Ascore','Cscore','Impulsive', 'SS','Alcohol','Amphet','Amyl','Benzos','Caff','Cannabis','Choc','Coke','Crack','Ecstasy','Heroin','Ketamine', 'Legalh','LSD','Meth','Mushrooms','Nicotine','Semer','VSA'))
drug_use <- drug_use %>% mutate_at(as.ordered, .vars=vars(Alcohol:VSA))
drug_use <- drug_use %>%
mutate(Gender = factor(Gender, labels=c("Male", "Female"))) %>%
mutate(Ethnicity = factor(Ethnicity, labels=c("Black", "Asian", "White",
"Mixed:White/Black", "Other",
"Mixed:White/Asian",
"Mixed:Black/Asian"))) %>%
mutate(Country = factor(Country, labels=c("Australia", "Canada", "New Zealand",
"Other", "Ireland", "UK", "USA")))
```

1(a).Define a new factor response variable recent_cannabis_use which is “Yes” if a person has used cannabis
within a year, and “No” otherwise. 
```{r}
drug_use <- drug_use %>% mutate(recent_cannabis_use = as.factor(ifelse(Cannabis < 'CL3', 'No', 'Yes')))
str(drug_use)
```
1(b)Split drug_use_subset into a training data set and a test data set called drug_use_train and drug_use_test.
The training data should include 1500 randomly sampled observation and the test data should include the
remaining observations in drug_use_subset.
```{r}
set.seed(1)
drug_use_subset <- drug_use %>% select(Age:SS, recent_cannabis_use)
train <- sample(1:nrow(drug_use_subset),1500)
drug_use_train <- drug_use_subset[train,]
drug_use_test <- drug_use_subset[-train,]
dim(drug_use_train)
dim(drug_use_test)
```
1(c) Fit a logistic regression to model recent_cannabis_use as a function of all other predictors in
drug_use_train. Fit this regression using the training data only.
```{r}
log_mod <- glm(recent_cannabis_use~. , data = drug_use_train , family = binomial)
summary(log_mod)
```
2.Decision Tree Model
```{r}
#a
set.seed(1)
tree_parameters <- tree.control(nobs=nrow(drug_use_train), minsize=10, mindev=1e-3)
tree_mod <- tree(recent_cannabis_use ~ . , control = tree_parameters, data = drug_use_train)
cvtree <- cv.tree(tree_mod, K = 10, FUN = prune.misclass)
best_size <- min(cvtree$size[which(cvtree$dev == min(cvtree$dev))])
best_tree <- prune.tree(tree_mod, best = best_size)

```
By drawing the tree, we can see that the first split of our tree is by the `Country` variable.
```{r}
#b
draw.tree(best_tree, nodeinfo = TRUE)
```

The TPR of our predictions is 0.7867299 and the FPR is 0.1666667.
As the *true positive rate* (TPR) is calculated by $\frac{TP}{TP+FN}$, we divide the bottom right element by the second column of our confusion matrix. Likewise, the *false positive rate* (FPR) is calculated by $\frac{FP}{FP+TN}$ which can be obtained by dividing the lower left element by the first column of our confusion matrix.
```{r}
pred.val <- predict(best_tree, drug_use_test, type = 'class')
pred.val1 <- predict(best_tree, drug_use_test, type = 'vector')
err <- table(pred.val, drug_use_test$recent_cannabis_use)
tpr <- err[2,2]/(err[2,2] + err[1,2])
fpr <- err[2,1]/(err[2,1] + err[1,1])
```

3.ROC curve for decision tree
```{r}
pred <- prediction(pred.val1[,2],drug_use_test$recent_cannabis_use)
perf <- performance(pred, measure = 'tpr',x.measure = 'fpr')
plot(perf, col = 2, lwd = 3, main = 'ROC Curve for decision tree')
abline(0,1)
```
ROC curve for logistic regression
```{r}
pred_log <- predict(log_mod, drug_use_test,type = 'response')
pred2 <- prediction(pred_log, drug_use_test$recent_cannabis_use)
perf2 <- performance(pred2, measure = 'tpr',x.measure = 'fpr')
plot(perf2, col = 2 , lwd = 3 , main = 'ROC curve for logistic regression')
abline(0,1)
```
3(b).Compute AUC for both models and print them. AUC for decision tree is 0.8348046 and AUC for logistic regression is 0.8792247. We can find AUC for logistic regression is larger.
```{r}
auc_tree <- performance(pred,'auc')@y.values
auc_log <- performance(pred2,'auc')@y.values
```
4(a)Convert type column to factor. Use the table command to print the number of patients with each leukemia subtype. Which leukemia subtype occurs the least in this data?
BCR-ABL occurs the least in this data.
```{r}
leukemia_data <- read_csv("leukemia_data.csv")
leukemia_data <- leukemia_data %>% mutate(Type = as.factor(leukemia_data$Type))
leukemia_data
table(leukemia_data$Type)
```
Run PCA on the leukemia data using prcomp function with scale=TRUE and center=TRUE (this scales
each gene to have mean 0 and variance 1).
```{r}
pr.out <- prcomp(leukemia_data %>% select(-Type), scale = TRUE, center = TRUE)
pr.var <- pr.out$sdev ^ 2
pve <- pr.var / sum(pr.var)
cumulative_pve <- cumsum(pve)
## This will put the next two plots side by side
par(mfrow=c(1, 2))
## Plot proportion of variance explained
plot(pve, type="l", lwd=3)
plot(cumulative_pve, type="l", lwd=3)
```
4c.Use the results of PCA to project the data into the first two principal component dimensions. prcomp
returns this dimension reduced data in the first columns of x. Plot the data as a scatter plot using plot
function with col=plot_colors where plot_colors is defined
```{r}
rainbow_colors <- rainbow(7)
plot_colors <- rainbow_colors[leukemia_data$Type]
new_coords <- pr.out$x[,1:2]
plot(new_coords, xlim = c(-60,60), ylim = c(-60,40), cex = 0 , main = 'PC1 & PC2')
text(-new_coords, label = leukemia_data$Type, cex = 0.3, col = plot_colors)
abline(h = 0, v = 0, col = 'lightblue', lty = 3)
```
T-ALL is most clearly separated from the others along the PC1 axis.
```{r}
head(sort(abs(pr.out$rotation[,1]),decreasing = TRUE))
```
4f.
```{r}
library(dendextend)
leukemia_subset <- leukemia_data %>% filter(Type == c('T-ALL', 'TEL-AML1', 'Hyperdip50'))
dis <- dist(leukemia_subset[,-1], method = 'euclidean')
leukemia.hc <- hclust(dis, method = 'complete')
dend1 <- as.dendrogram(leukemia.hc)
dend1 <- color_branches(dend1, k = 3)
dend1 <- color_labels(dend1, k = 3)
dend1 <- set(dend1, 'labels_cex', 0.3)
dend1 <- set_labels(dend1, labels = leukemia_subset$Type[order.dendrogram(dend1)])
plot(dend1, horiz = T, main = 'Dendrogram colored by three clusters')

dend2 <- as.dendrogram(leukemia.hc)
dend2 <- color_branches(dend2, k = 5)
dend2 <- color_labels(dend2, k = 5)
dend2 <- set(dend2, 'labels_cex', 0.3)
dend2 <- set_labels(dend2, labels = leukemia_subset$Type[order.dendrogram(dend2)])
plot(dend2, horiz = T, main = 'Dendrogram colored by five clusters')
```
